Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-05-01 demystify.tex
updates

Finally, for fellowships (\autoref{sec:fellowships}) you might get paid over the summer depending on your fellowship (\autoref{sec:fellowships}). Typically good ones, e.g., from NSF, Google, Microsoft, will pay you the whole year.

@@

\paragraph{Ranks} Faculty have ranks such as assistant, associate, and full, regardless if they are tenure-line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. A tenure track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3-4 years, some 10+ years, some remain associate). A non-tenure track faculty, e.g., teaching or research faculty, also starts with assistant, then is promoted to associated, and full prof.

\subsection{Tenured or tenure-track faculty? Who do you choose?}\label{sec:tenure-vs-tenure-track}

@@

The short answer is that tenure-track faculty such as assistant professors are likely young and active in research (they have to, to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding (yet).

Tenured faculty, e.g., associate and full profs., are likely older, more well-known, and have more experience in managing students.  However, they might not push you as hard. They might also expect you to figure things out yourself, i.e., you need to be independent.  Some tenured faculty are also no longer active in research and are more involved with administrative responsibilities or with their startup companies (this means they will likely not take new students).

@@

First, read the prof's website, and see if they say something about contacting them. Many profs. explicitly indicate how prospective students should (or should not) contact them (e.g., using specific email subjects).
In general, the best way to catch the prof.'s attention is to \emph{customize your email} for them.  For example, read their papers, know what they work on, and see if you are interested in their research. Then email them and talk about how/why their work would match yours.
In contrast, if you write a generic email that can be sent to multiple professors (e.g., if you just change some names and keywords in the email or copy and paste paper titles), you will not get a response.

@@

However, if your goal is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. A non-CS faculty can serve in the PhD dissertation committee (common) or \emph{co-advise} (somewhat rare), but your main PhD adviser will likely be a tenure-line faculty in CS (\autoref{sec:faculty-types}). 
For example, a prof. in Stats or Math might be able to serve as a co-adviser, but not as a sole adviser of a student in a CS PhD program.  Similarly, a prof. in CS cannot be the sole adviser and graduate a PhD student in Math or Stats.
If in doubt, check with the CS department for their requirements.

@@

\paragraph{Low English exam scores (e.g., IELTS or TOEFL)}  While profs. might not care much about these, there might be a certain minimum set by the college or the university that you need to pass to get admitted, especially for TA funding (e.g., you need to communicate well with students as a TA).  Thus, while professors are willing to argue for your case, they might be reluctant to go against the standard requirement of the university on English scores.  Note this is for international students, as domestic ones (or those from these \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{countries}) do not have to take these tests.

\paragraph{Overqualified or Lack Interest}  This might be paradoxical and is the opposite of aiming too high. However if adcom believes that you are likely to get admitted and go to a better program (i.e., they are just your ``safety'' schools), they might not be so enthusiastic to admit you and want to save the spot for someone else.

@@

Understandably students get excited about their upcoming PhD journey and want to buy new computer equipment and electronics to prepare. However, you should check with your adviser first.  They might have funding to buy you computers and other equipment (e.g., software, books, keyboards, headphones, tablets, etc). Many CS programs also provide some budget or computer equipment to incoming PhD students (e.g., a laptop).

>>>>

Finally, for fellowships (\autoref{sec:fellowships}) you might get paid over the summer depending on your fellowship (\autoref{sec:fellowships}). Good ones, e.g., from NSF, Google, and Microsoft, will pay you the whole year.

@@

\paragraph{Ranks} Faculty have ranks such as assistant, associate, and full, regardless if they are tenure-line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. Typically, a tenure-track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3--4 years, some 10+ years, some remain associate). A non-tenure track faculty, e.g., teaching or research faculty, also starts with assistant, then is promoted to associated, and full prof.

\subsection{Tenured or tenure-track faculty?}\label{sec:tenure-vs-tenure-track}

@@

Tenure-track faculty such as assistant professors are likely young and very active in research (they have to, to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding (yet).

Tenured faculty, e.g., associate and full profs., are likely older, more well-known, and have more experience in managing students.  However, they might not push you as hard and expect you to figure things out yourself, i.e., you need to be independent.  Some tenured faculty are also no longer active in research and are more involved with administrative responsibilities or with their startup companies (this means they will likely not take new students). \autoref{sec:faculty-types} gives more details on faculty types.

@@

First, read the prof's website, and see if they say something about contacting them. Many profs. explicitly state how prospective students should (or should not) contact them, e.g., using specific email subjects.
In general, the best way to catch the prof.'s attention is to \emph{customize your email} for them.  For example, read their papers, know what they work on, and see if you are interested in their research. Then email them and talk about how/why their work would match yours.
In contrast, if you write a generic email that can be sent to multiple professors (e.g., if you just change some names and keywords in the email or copy and paste paper titles), you will not get a response.

@@

However, if your goal is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. A non-CS faculty can serve in the PhD dissertation committee (common) or \emph{co-advise} (somewhat rare), but your main PhD adviser will likely be a tenure-line faculty in CS (\autoref{sec:faculty-types}). 
For example, a prof. in Stats or Math might be able to serve as a co-adviser, but not as a sole adviser of a student in a CS PhD program. 
If in doubt, check with the CS department for their requirements.

@@

\paragraph{Low English exam scores (e.g., IELTS or TOEFL)}  While profs. might not care much about these, there might be a certain minimum set by the college or the university that you need to pass to get admitted, especially for TA funding (e.g., you need to communicate well with students as a TA).  Thus, while profs. are willing to argue for your case, they might be reluctant to go against the requirement of the university on standard exams.

Note this is for international students, as domestic ones (and likely those from these \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{countries}) do not have to take these tests.

\paragraph{Overqualified or Lack Interest}  This might be paradoxical and is the opposite of aiming too high. However if adcom believes that you are likely to get admitted and go to a better program, they might not be so enthusiastic to admit you and want to save the spot for someone else.

@@

Understandably students get excited about their upcoming PhD journey and want to buy new computer equipment and electronics to prepare. However, you should check with your prof. first.  They might have funding to buy you computers and other equipment (e.g., software, books, keyboards, headphones, tablets, etc). Many CS programs also provide budget or computer equipment to incoming PhD students (e.g., a laptop).
